Data Preparation

### 2021-07-21 x0_data-preparation.Rmd
geo coding

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("ggnewscale")

library(tidyverse)
library(lubridate)
library(zoo)
library(patchwork)
library(ggnewscale)

@@

  fb <- fb %>% 
  transmute( ADEP = origin
            ,ADES = destination
            ,TYPE = as.factor(typecode)
            ,DATE = date(day)
            ,CALL = callsign
            #,ACFT_ID = aircraft_uid    # throws an error in today's data download
            )

@@

prep_base_data <- function(.fn = file_names, ...){
ds <- read_osn_covid(.fn)
ds <- ds %>% append_ad_ctry()
ds <- ds %>% append_region()

@@

ds <- ds %>% drop_na()  # drop all ~ assumed to be clean

@@

loop deactivated as base data has been saved out and shared across the project team

```{r, eval=FALSE}
# read all files (note adpat function name)
fns <- list_apt_files()

@@

```{r, eval=TRUE, message=FALSE}
#quick peek

@@

ds1 %>% ggplot(aes(x = DATE)) +
  geom_line(aes(y = THRESHOLD)) +
  geom_point(aes(y = N, color = N > THRESHOLD))
```

```{r}
#CHECK WEEKLY DISTRIBUTION
ds2 <- ds1 %>% 
  mutate( WEEK_DAY = wday(DATE, label = TRUE)
         ,W_END    = WEEK_DAY %in% c("Sat", "Sun"))
uplim <- .95
lowlim <- .20

@@

ds2 %>% ggplot(aes(x = DATE)) +
  geom_col(aes(y = MED, fill = !(MED < LOW_THR | MED > UP_THR))) +
  geom_line(aes(y = UP_THR), color = "blue") +
  geom_line(aes(y = LOW_THR), color = "red") +
  scale_x_date(limits = c(as.Date("2019-10-01"), as.Date("2020-12-01")), date_breaks = "30 days", date_labels = "%b\n%y") +
  labs(y = "7-day rolling median", x = NULL, fill = "Normal operation?",
       title = "What is normal?",
       subtitle = "Red line - Lower threshold; Blue line - upper threshold",
       caption = "Lower threshold set to 20th percentile of the last 180 days;\nUpper threshold set to 95th percentile of the last 180 days"
)
```

```
# breadcrumb ------------- need to check what this is for --------------------
#ds2 %>% ggplot() +
  geom_freqpoly(aes(x = N, color = W_END))
```

```{r}
us <- full_dataset %>% 
  filter(ADEP_REG == "US" | ADES_REG == "US") %>%
  group_by(YR = year(DATE), DATE) %>% 
  summarize(N = n(), .groups = "drop")

# static thresholds, x-th percentile rounded to lower integer
pcts   <- c(0.8, 0.6, 0.3, 0.1)
us_ref <- us %>% filter(YR == 2019) %>%
  pull(N) %>% quantile(probs = pcts) %>% floor()

us_ref2 <- us %>% filter(YR == 2019, DATE >= lubridate::ymd("2019-07-01")) %>%
  pull(N) %>% quantile(probs = pcts) %>% floor()

max_N <- max(us$N)
lvls  <- c("First", "Second", "Third", "Fourth", "Fifth")
us <- us %>% 
  mutate( THRES  = cut(N, breaks = c(max_N, us_ref, 0),  labels = lvls)
         ,THRES2 = cut(N, breaks = c(max_N, us_ref2, 0), labels = lvls)
         )
```
```{r}
us %>%
  ggplot(aes(x = DATE)) +
 # geom_line(aes(y = THRES)) +
  geom_point(aes(y = N, color = THRES2)) +
  geom_line(aes(y = N, colour = THRES2, group = 1)) +
  geom_hline(yintercept = c(max_N, us_ref2, 0))
```

```{r}
# comparator levels
uplim  <- .80
lowlim <- .20
slide  <-  60

tmp <- us %>% 
  mutate( N_sm    = roll_quantile(N, width = 7, p = .50)
         ,MED     = roll_quantile(N, width = 14, p = .50, min_obs = 7)
         ,UP_THR  = roll_quantile(N, width = slide, p = uplim,  min_obs = 7)
         ,LOW_THR = roll_quantile(N, width = slide, p = lowlim, min_obs = 7)
         ,LOW_CRIT= LOW_THR >= MED
         ,UP_CRIT = UP_THR  <= MED
         ,STATE   = case_when( !LOW_CRIT & !UP_CRIT ~ "LEVEL"
                              , LOW_CRIT & !UP_CRIT ~ "CRUNCH"
                              ,!LOW_CRIT &  UP_CRIT ~ "RECOV" 
                              ,TRUE ~ "UNKWN")
         ,BAND    = 4000
         ) %>%
  filter(DATE >= lubridate::ymd("2019-10-01"))

###---------Rainer, the "new_scale_colour()" kinda resets the scale, and you can use another one for the next geom_lines. In the example below, I left the legend.position = "bottom" (instead of "NULL"), just to see the colors------###
tmp %>%
  ggplot() +
  geom_line(aes(x = DATE, y = N), colour = "grey80") +
  geom_line(aes(x = DATE, y = N_sm, colour = THRES2, group = "RQ"), size = 2) +
  geom_line(aes(x = DATE, y = UP_THR), color = "blue") +
  geom_line(aes(x = DATE, y = LOW_THR), color = "red") +
  geom_line(aes(x = DATE, y = MED), color = "green") +
  scale_color_brewer(type = "seq") +
  new_scale_colour() +
  geom_line(aes(x = DATE, y = BAND, group= "1"), colour = if_else(tmp$LOW_CRIT, "red","green"), size = 10) +
  geom_line(aes(x = DATE, y = BAND-500, group= "1"), colour = if_else(tmp$UP_CRIT, "red","green"), size = 4) +
  geom_line(aes(x = DATE, y = BAND-1500, group= "1", colour = STATE), size = 4) +
  scale_colour_brewer(type = "qual") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Airport Level Considerations

top-5 in the whole sample, confirms preference :)

```{r}
full_dataset %>% 
  group_by(ADEP_REG, ADEP) %>% 
  summarise(N = n()) %>% 
  slice_max(order_by = N, n = 5)
```

```{r}
sbgr <- full_dataset %>% filter(ADEP == "SBGR")
```

```{r}
apt_timeline <- function(.df){
  df <- .df %>% group_by(DATE) %>%
    summarise(N_DEP = n()
              ,DEP_DOM = sum(ADEP_CTRY == ADES_CTRY)
              ,DEP_REG = sum(ADEP_REG == ADES_REG))
}

plot_timeline <- function(.df){
  df <- .df %>% 
  mutate(N_DEP_SM = roll_quantile(N_DEP, width = 7, p = .50))

  p <- df %>%  
  ggplot() +
  geom_line(aes(x = DATE, y = N_DEP), colour = "grey60") +
  geom_line(aes(x = DATE, y = N_DEP_SM)) +
  scale_x_date( limits = c(as.Date("2019-10-01"), as.Date("2021-06-01"))
               ,date_breaks = "60 days", date_labels = "%b\n%y") +
  theme_minimal()

  return(p)
}

sbgr %>% apt_timeline() %>% plot_timeline()
```

```{r}
sbgr_tl %>%
  ggplot() +
  geom_point(aes(x = DATE, y = DEP_DOM / N_DEP)) +
  theme_minimal()
```
```{r}
dep_vs_domshare <- function(.df){
  p <- .df %>%
  ggplot() +
  geom_point(aes(x = N_DEP, y = DEP_REG / N_DEP, colour = as.factor(year(DATE)))) +
  labs(colour = NULL) + 
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)
                     ,expand = c(0,0)) +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.8))
return(p)
}

sbgr_tl %>% dep_vs_domshare()
```

```{r}
check_apt <- function(.mvts, .apt){
  apt_mvts <- .mvts %>%
    filter(ADEP == .apt) %>% 
    apt_timeline()

  p1 <- apt_mvts  %>% plot_timeline()
  p2 <- apt_mvts %>% dep_vs_domshare()

  p <- (p1 + labs(subtitle = .apt)) + p2
return(p)
}

full_dataset %>% check_apt("EGLL")
```

```{r}
full_dataset %>% check_apt("EDDF")
```

```{r}
full_dataset %>% check_apt("LFPG")
```

```{r}
full_dataset %>% check_apt("KATL")
```

```{r}
full_dataset %>% check_apt("KORD")
```

```{r}
full_dataset %>% check_apt("KJFK")
```

```{r}
full_dataset %>% check_apt("EHAM")
```

```{r}
full_dataset %>% check_apt("EHAM")
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(zoo)
library(sf)
library(countrycode)

@@

  fb <- fb %>% 
  transmute( ADEP = origin
            ,ADES = destination
            ,ICAO24 = icao24
            ,TYPE = as.factor(typecode)
            ,DATE = date(day)
            ,CALL = callsign
            #,ACFT_ID = aircraft_uid    # throws an error in today's data download
            ,LAT1 = latitude_1
            ,LON1 = longitude_1
            ,ALT1 = altitude_1
            ,LAT2 = latitude_2
            ,LON2 = longitude_2
            ,ALT2 = altitude_2
            )

@@

# ====================== geo coding stuff ===================================

#' utility function to geocode lat lon positions - read shape files for geocoding
#' 
load_shapefile_for_geo_ref <- function(){
  EEZ_land_shp <- list.files("../__DATA/xWorld_EEZ_and_land_v3", pattern = ".shp$", full.names = TRUE)
  eez_land <- sf::read_sf(EEZ_land_shp) %>% select(ISO_TER1)
}

#' utility function to coerce

# ====================== wrap all prep in one function ===================

prep_base_data <- function(.fn = file_names, ...){
ds <- read_osn_covid(.fn)
ds <- ds %>% append_ad_ctry()

ds_ok  <- ds %>% filter(!is.na(ADEP), !is.na(ADES))
ds_nok <- ds %>% filter( is.na(ADEP) | is.na(ADES))

ds <- ds %>% append_region()

@@

#ds <- ds %>% drop_na()  # drop all ~ assumed to be clean

@@

```{r}
# read all files (note adpat function name)
fns <- list_apt_files()

@@

```{r, eval=TRUE}
#quick peek

@@

ds1 %>% ggplot(aes(x = DATE)) +
  geom_line(aes(y = THRESHOLD)) +
  geom_point(aes(y = N, color = N > THRESHOLD))

#CHECK WEEKLY DISTRIBUTION
ds2 <- ds1 %>% mutate(WEEK_DAY = wday(DATE, label = TRUE), W_END = WEEK_DAY %in% c("Sat", "Sun"))
uplim <- .95
lowlim <- .20

@@

ds2 %>% ggplot(aes(x = DATE)) +
  geom_col(aes(y = MED, fill = !(MED < LOW_THR | MED > UP_THR))) +
  geom_line(aes(y = UP_THR), color = "blue") +
  geom_line(aes(y = LOW_THR), color = "red") +
  scale_x_date(limits = c(as.Date("2019-10-01"), as.Date("2020-12-01")), date_breaks = "30 days", date_labels = "%b\n%y") +
  labs(y = "7-day rolling median", x = NULL, fill = "Normal operation?",
       title = "What is normal?",
       subtitle = "Red line - Lower threshold; Blue line - upper threshold",
       caption = "Lower threshold set to 20th percentile of the last 180 days;\nUpper threshold set to 95th percentile of the last 180 days"
)

#ds2 %>% ggplot() +
  geom_freqpoly(aes(x = N, color = W_END))
```
